---
title: "Fit HHMMs: What if the TPM covariates were (more) nonlinear?"
format: html
editor: source
---


```{r setup, include=FALSE}
library(tidyverse)
library(FBtagtools)
library(momentuHMM)
knitr::opts_chunk$set(echo = TRUE, cache = FALSE,
                      warning = FALSE)

# if this is round 2+ of fitting the HHMMs
repeating <- TRUE
test_run <- TRUE # fastest/least number of iterations for code debugging
```

```{r}
source('../utils/getLL.R')
```

# Read in Data and Models

## Tag Data

```{r, data-in}
# note that for the public repo, data is stored as part of the fitted model objects
# so if you want to run this code, comment the code below and get the data there, e.g.
# hhmm_data <- readRDS("../data/fitted-models/sel_both.RDS")
# hhmm_data <- hhmm_data$data

if (sum(grepl("Mac", Sys.info()))){
  ncpath <- "/Users/sld33/Dropbox/FBdata"
}else{
  ncpath <- "C:/Users/stacy/Dropbox/FBdata"
}

hhmm_data <- readRDS(file.path(ncpath, 'hhmm_smrt_data_aug_2025.RDS'))
```

```{r}
#| label: prep-hhmm-data

hhmm_data <- hhmm_data |>
  prepData(coordNames = NULL,
           covNames = c(
             "mfas_presence",
             "mfa_bb_rms_max_scale",
             "mfa_csel_scale",
             "sonar_before_eoc",
             "fine_mfa_bb_rms_max_scale",
             "fine_mfa_csel_scale"),
           hierLevels = c("1", "2i", "2"))
```


# Setup for HHMM

<!-- code to set up hierarchical structure and design matrix for the HHMMs is in the file _hhmm-setup.qmd. -->

{{< include _hhmm-setup.qmd >}}

## Starting values

Initially, use starting values from coarse-only or fine-only HMMs.

If re-fitting a new round of models (if `repeating` is `TRUE`), then start with the best models fitted so far.

Users taking this code from the repository and changing nothing will have `repeating = TRUE` and will re-fit models based on the best fitted models reported in the paper.

```{r}
if (repeating){
model_files <- dir(path = "../data/fitted-models/",
                   pattern = ".RDS")
            
model_meta <- 
  data.frame(filename = paste0("../data/fitted-models/",
                               model_files)) |>
  mutate(
    new_model_file = str_replace_all(filename, pattern = "fitted", 
                                     replacement = "refitted")) |>
    # need to remove the "../data..."
    separate_wider_delim(filename, delim = "/", 
                                      names = c(NA, NA, NA, "model_name"),
                         cols_remove = FALSE) |>
    mutate(model_name = str_remove_all(model_name, 
                                pattern = fixed(".RDS"))
    ) |>
  as.data.frame()

start_models <- list()
for (m in c(1:nrow(model_meta))){
  start_models[[m]] <- readRDS(model_meta[m, "filename"])
  model_meta$full_formula[m] <- as.character(start_models[[m]]$conditions$formula)[2]
}

# of course we could not set up the formulas this way if the models were not prefitted.
# but this is a way to get a list of all options included in the paper w/o hand typing them
model_meta <- model_meta |>
  separate_wider_delim(full_formula, delim = "I((level == \"2i\") * 1)",
                       names = c("coarse_formula", "fine_formula"),
                       cols_remove = FALSE) |>
  mutate(coarse_formula = str_remove_all(coarse_formula, 
                                         pattern = fixed("I((level == \"1\") * ")) |>
           str_squish(),
         coarse_formula = str_remove_all(coarse_formula, fixed(")")),
         coarse_formula = str_remove_all(coarse_formula, fixed("0 + 1 +")),
         coarse_formula = ifelse(str_ends(coarse_formula, fixed("+")),
                                 word(coarse_formula, start = 1, end = -2),
                                 coarse_formula),
         fine_formula = str_remove_all(fine_formula, fixed('+ I((level == "2") * 1)')),
         fine_formula = str_remove_all(fine_formula, fixed('+ I((level == "2") * ')),
         fine_formula = str_remove_all(fine_formula, fixed(")")),
         fine_formula = if_else(is.na(fine_formula), "1", fine_formula),
         fine_formula = paste0("1", ifelse(fine_formula %in% c("", " "), "", "+"), 
                               str_squish(fine_formula)),
         coarse_formula = paste0("1", ifelse(coarse_formula %in% c("", " "), "", "+"),
                                 str_squish(coarse_formula))
         ) |>
  data.frame()

} # end of "if repeating..."

spline_model_meta <- model_meta[model_meta$model_name == "sel_both",] |>
  mutate(model_name = paste0("spline_", model_name),
         new_model_file = paste0("../data/refitted-models/", model_name, ".RDS"),
         coarse_formula = str_replace_all(coarse_formula,
                                          pattern = "mfa_csel_scale",
                                          replacement = "ns(mfa_csel_scale, Boundary.knots =c(0,1), knots = c(0.2, 0.4, 0.6, 0.8))")) |>
  data.frame()
```

# Fit Models


```{r, refit-models}
n_fit_cycles <- ifelse(test_run, 1, 10)

  # if at least one round of these fits is already done, 
  # start w/previously fitted editions
  fitted_models <- list()
  fitted_models[[1]] <- start_models[[which(model_meta$model_name == "sel_both")]]


these_fitted_models <- list()

for (c in c(1:n_fit_cycles)){
  for(rep in c(1:2)){
    for (m in c(1:nrow(spline_model_meta))){
      message(paste('Round', c, 'of', n_fit_cycles, ', rep ', rep, 
                    '- Fitting model', m, 'of', nrow(spline_model_meta), 
                    '. Model name:', spline_model_meta[m, "model_name"]))
      

        hierFormula <- data.tree::Node$new("Ziphius HHMM TPM covariates")
        hierFormula$AddChild(name = "level1", 
                             formula = as.formula(paste0("~", spline_model_meta[m, "coarse_formula"])))
        hierFormula$AddChild(name = "level2", 
                             formula = as.formula(paste0("~", spline_model_meta[m, "fine_formula"])))
        fixPar <- data.tree::Node$new("Ziphius HHMM fixed values")
        fixPar$AddChild(name = "beta")
        fixPar$beta$AddChild(name = "level2")
        my_fixies <- fitted_models[[m]]$conditions$fixPar$beta
        if (str_detect(model_meta[m, "model_name"], pattern = "fine") |
            str_detect(model_meta[m, "model_name"], pattern = "both")){
          which_fix <- str_starts(colnames(my_fixies), pattern = "1|2|3|4") &
            str_ends(colnames(my_fixies), pattern = "1|2|3|4")
          which_rows <- str_detect(rownames(my_fixies), pattern = fixed('level == "2"'))
          my_fixies[nrow(my_fixies), which_fix] <- 0
          my_fixies <- my_fixies[which_rows,which_fix]
          fixPar$beta$level2$AddChild(name = "dive_cycle",
                                      fixPar = my_fixies)
      }
      
     
        if (rep == 1){
          ## CODE ERRORS HERE: can't get hierFormula processed with the ns() in it.
          ## MAY NEED TO UPDATE MOMENTUHMM TO TRY THIS.
          Par0 <- getPar0(fitted_models[[m]],
                          hierFormula = hierFormula)
          hierBeta0 <- Par0$hierBeta
          hierDelta0 <- Par0$hierDelta
          Par0 <- Par0$Par
        }else{
          Par0 <- getPar0(fitted_models[[which.max(unlist(lapply(fitted_models, getLL)))]],
                          hierFormula = hierFormula)
          hierBeta0 <- Par0$hierBeta
          hierDelta0 <- Par0$hierDelta
          Par0 <- Par0$Par
        }
      }
      
      # fit model
      this_file <- spline_model_meta[m, 'new_model_file']

      this_fit <- try(fitHMM(hhmm_data,
                             hierStates = hierStates,
                             hierDist = hierDist,
                             hierFormula = hierFormula,
                             Par0 = Par0,
                             DM = DM,
                             hierBeta = hierBeta0,
                             hierDelta = hierDelta0,
                             optMethod = "nlm",
                             retryFits = ifelse(test_run, 1, 10),
                             retrySD = ifelse(rep == 1, 25, 10)
      ),
      silent = TRUE
      )
      if ("momentuHMM" %in% class(this_fit)){
        these_fitted_models[[m]] <- this_fit
        if (c == 1 & rep == 1 & !repeating){
          # first time thru always save the result
          fitted_models[[m]] <- this_fit
          saveRDS(fitted_models[[m]], this_file)
        }
        # else only save if it's better than what we already have
        if (getLL(this_fit) > max(-Inf, getLL(fitted_models[[m]]), na.rm = TRUE)){
          fitted_models[[m]] <- this_fit
          saveRDS(fitted_models[[m]], this_file)  
        }
      }
    }
  }# end of loop over models
} # end fitting cycles
```
