

# Likelihood Weighting Simulation Study

## Methods

Facing challenges in identifying appropriate starting values for all HHMM model parameters and worried that dive-cycle-level observations might be "swamped" by the inevitably much more numerous 5-minute-scale observations in the model fitting process, we explored the idea of weighting the model likelihood by time-step duration.

We carried out a small simulation study to assess the effectiveness of adding weights on the likelihood function such that the likelihood of each coarse-scale time-step was multiplied by the median number of five-minute intervals per dive cycle across the full dataset. 

We chose not to weight dive cycles by the exact observed number of five-minute periods they contained since that strategy, while logical, would afford greatest weight to the longest dives cycles. Since previous work has indicated that *Ziphius* perform longer dive cycles in response to MFAS exposure, we chose equal weighting of all dive cycles as a conservative approach. 

We simulated from two models: HHMMs like those fitted in this study, with and without an effect of MFAS RMS RL on the coarse-scale transition probability matrix. For each of the 100 simulations for each model, we generated a dataset the same size as the observed dataset (13 individuals and a total of 569 dive cycles), and fitted four HHMMs: with and without MFAS covariate, and with and without likelihood weights. We determined confidence interval coverage and model selection accuracy in each scenario. 


## Results

Models that were fitted employing weighting of the likelihood function (to correspond to the longer temporal duration of the coarse-scale states) did not perform well: in sets of 100 simulations, 95% confidence intervals for the parameter measuring the MFAS effect on the probability of transition between coarse-scale states had coverage of just 27%, compared to 96% for the unweighted models (Table S9, Figure S42). 

On the other hand, the weighted models had 99% power to detect MFAS effects that were truly present, compared to 75% for the unweighted models. 

When the true model included no MFAS effect, the unweighted model falsely detected it only 1% of the time, compared to 88% for the weighted model (Table S11). 

All these results were driven by underestimation of uncertainty in weighted models, yielding unrealistically small standard errors and narrow confidence intervals. 

**Because weighting would result in misleading inference, all models presented in the main body of the paper were fitted without it.** Simulation results are included in these supplemental materials mainly as a caution to anyone considering a similar approach.


### Case 1: True Model Includes MFAS Effect

<!-- ### Read in Results -->

```{r}
sim_files <- list.files(path = "../data/simulations/sonar",
                        pattern = "*.RDS")

sim_results <- list()

for (f in c(1:length(sim_files))){
  sim_results[[f]] <- bind_rows(
    readRDS(
      file.path(
        "..",
        "data",
        "simulations",
        "sonar",
        sim_files[f]))) |>
    mutate(original_file = f)
}

sim_results <- bind_rows(sim_results) |>
  drop_na(LL)

sim_CI <- sim_results
```

#### AIC and likelihood results

```{r}
sim_results <- sim_results |>
  mutate(weighted = if_else(str_detect(model_name, "_wt"),
                            "Yes",
                            "No"),
         model_name = str_remove_all(model_name, "_wt")) |>
  pivot_wider(names_from = "model_name",
              values_from = c("AIC", "LL"),
              id_cols = c("sim_num", "original_file", "weighted")) |>
  mutate(dAIC = AIC_no_sonar - AIC_rl_coarse,
         dLL =  LL_rl_coarse - LL_no_sonar)
```

::: {#suppfig-aic-with-weights}

```{r}
gf_violin(dAIC ~ weighted, data = sim_results) |>
  gf_sina() |>
  gf_refine(scale_y_continuous(trans='log2')) |>
  gf_labs(y = "AIC difference\n(with minus without MFAS covariate)",
          x = "Weighted Likelihood?")
```

By what margin would model selection using AIC select the true model (with MFAS effect), with and without duration-weighting applied to the likelihood?
:::

::: {#supptab-daic-with-weights}

```{r}

selection_table1 <- mosaic::prop(~dAIC >= 6 | weighted, data = sim_results) |>
  data.frame()
names(selection_table1) <- "Proportion dAIC > 6"

selection_table1 <- selection_table1 |>
  mutate(`Weighting?` = c("No", "Yes")) |>
  select("Weighting?", "Proportion dAIC > 6")
  
  gt(selection_table1)
```

Proportion of simulations where the difference in AIC (model with MFAS minus model without) exceeded 6, when the true model was one with MFAS effect.
:::

#### Confidence Interval Coverage


```{r}
fitted_model <- 
  readRDS("../data/simulations/sim_starting_model.RDS")
```

```{r}  
true_params <- fitted_model$mle 
true_mfas <- true_params$beta['I((level == "1") * mfa_bb_rms_max_scale)',
                      c("1 -> 5", "5 -> 1")]
```

```{r}
sim_CI <- sim_CI |>
  mutate(true_15 = true_mfas["1 -> 5"],
         true_51 = true_mfas["5 -> 1"]) |>
  mutate(true_15_in_CI = if_else(t15lower <= true_15 &
                                   t15upper >= true_15,
                                 1,
                                 0),
         true_51_in_CI = if_else(t51lower <= true_51 &
                                   t51upper >= true_51,
                                 1,
                                 0),
         AE_15 = abs(t15est - true_15),
         AE_51 = abs(t51est - true_51),
         row_num = c(1:n()),
         pretty_name = case_when(model_name == "rl_coarse" ~ "No Weighting",
                                 model_name == "rl_coarse_wt" ~ "Weighted"))

CI_summary <- sim_CI |>
  drop_na(t15est) |>
  group_by(model_name) |>
  summarise(
    # n_NA_15 = sum(is.na(true_15_in_CI)),
    # n_NA_51 = sum(is.na(true_51_in_CI)),
    n_total = n(),
    prop_in_CI_15 = sum(true_15_in_CI, na.rm = TRUE) / sum(!is.na(true_15_in_CI)) * 100,
    prop_in_CI_51 = sum(true_51_in_CI, na.rm = TRUE) / sum(!is.na(true_51_in_CI)) * 100,
    MAE_15 = mean(AE_15, na.rm = TRUE) / abs(first(true_15)) * 100,
    MAE_51 = mean(AE_51, na.rm = TRUE) / abs(first(true_51)) * 100) |>
  ungroup() |>
  mutate(across((starts_with("MAE") | starts_with("prop_")), function(x) signif(x, 3))) |>
  rename(`Model Name` = model_name,
         # `NAs in P(T->V)` = n_NA_15,
         # `NAs in P(V->T)` = n_NA_51, # very hard to estimate b/c sample size too small
         `N Simulations` = n_total,
         `CI Coverage (%; P(T->V))` = prop_in_CI_15,
         `CI Coverage (%; P(V->T))` = prop_in_CI_51,
         `Mean Absolute Error (%; P(T->V))` = MAE_15,
         `Mean Absolute Error (%; P(V->T))` = MAE_51
         )
```

::: {#supptab-ci-summary}

```{r}
gt(CI_summary) |>
  tab_style(
    style = cell_text(size = pct(65)),
    locations = list(cells_body(), cells_column_labels()) 
  )
```

Confidence interval coverage in simulations, for simulations in which the true model included an effect of MFAS on the transition rates. "T->V" is the probability of transition from Typical to Variant state, and "V->T" is Variant to Typical.
:::

::: {#suppfig-mfas-cis-15}

```{r}
#| label: CIs-15
gf_point(t15est ~ row_num, 
         data = sim_CI |> filter(grepl(model_name, pattern = "rl") &
                                         !is.na(true_15_in_CI)),
         size = 0.2,
         color = ~factor(true_15_in_CI)) |>
  gf_theme(scale_color_manual(values = c("goldenrod", "black"))) |>
  gf_errorbar(t15lower + t15upper ~ row_num,
              color = "grey44", alpha = 0.5) |>
  gf_facet_wrap(~pretty_name, ncol = 1) |>
  gf_hline(yintercept = sim_CI[1, "true_15"], color = 'darkred', linetype = "dashed") |>
  gf_theme(legend.position = "none") |>
  gf_labs(y = "MFAS Parameter\n(Typical to Variant)",
          x = "Simulation Number")
```

Parameter estimates and 95% confidence intervals for the parameter measuring the effect of MFAS on transition from Typical to Variant state. Dotted horizontal line shown the true value of the parameter.
:::

\newpage

::: {#suppfig-mfas-cis-51}

```{r}
#| label: CIs-51
gf_point(t51est ~ row_num, 
         data = sim_CI |> filter(grepl(model_name, pattern = "rl") &
                                   !is.na(true_51_in_CI)),
         size = 0.2,
         color = ~factor(true_51_in_CI)) |>
  gf_theme(scale_color_manual(values = c("goldenrod", "black"))) |>
  gf_errorbar(t51lower + t51upper ~ row_num,
              color = "grey44", alpha = 0.5) |>
  gf_facet_wrap(~pretty_name, ncol = 1) |>
  gf_hline(yintercept = sim_CI[1, "true_51"], color = 'darkred', linetype = "dashed") |>
  gf_theme(legend.position = "none") |>
  gf_labs(y = "MFAS Parameter\n(Variant to Typical)",
          x = "Simulation Number")
```

Parameter estimates and 95% confidence intervals for the parameter measuring the effect of MFAS on transition from Variant to Typical state (note, this one is not estimated well because of the small size of the dataset). Dotted horizontal line shown the true value of the parameter.
:::


#### Power

Based on CIs, how often would we conclude there IS an MFAS effect (when there really is)?

::: {#supptab-power}

```{r}
A <- sim_CI |>
  group_by(model_name) |>
  drop_na(t15upper, t15lower) |>
  summarise(
    prop_MFAS_effect_15 = sum( 
      (sign(t15upper) * sign(t15lower)) == 1 &
        sign(true_15) == sign(t15upper)
    ) / n() * 100
  )

B <- sim_CI |>
  group_by(model_name) |>
  drop_na(t51upper, t51lower) |>
  summarise(
    prop_MFAS_effect_51 = sum( 
      (sign(t51upper) * sign(t51lower)) == 1 &
        sign(true_51) == sign(t51upper)
    ) / n() * 100
  )

left_join(A,B, by = join_by(model_name)) |>
  rename(`Model Name` = model_name,
         `Power (%, P(T->V))` = prop_MFAS_effect_15,
         `Power (%, P(V->T))` = prop_MFAS_effect_51
         ) |>
  mutate(across(where(is.numeric), function(x) signif(x, 3))) |>
  gt()
```

Power to detect MFAS effect when it is really present.
:::

### Case 2: True Model Excludes MFAS Effect

<!-- ### Read in Results -->

```{r}
sim_files2 <- list.files(path = "../data/simulations/no-sonar",
                        pattern = "*.RDS")

sim_results2 <- list()

for (f in c(1:length(sim_files2))){
  sim_results2[[f]] <- bind_rows(
    readRDS(
      file.path(
        "..",
        "data",
        "simulations",
        "no-sonar",
        sim_files2[f]))) |>
    mutate(original_file = f)
}

sim_results2 <- bind_rows(sim_results2) |>
 drop_na(LL)

sim_CI2 <- sim_results2
```

#### AIC and likelihood results

::: {#suppfig-dAIC2}

```{r}
sim_results2 <- sim_results2 |>
  mutate(weighted = if_else(str_detect(model_name, "_wt"),
                            "Yes",
                            "No"),
         model_name = str_remove_all(model_name, "_wt")) |>
  pivot_wider(names_from = "model_name",
              values_from = c("AIC", "LL"),
              id_cols = c("sim_num", "original_file", "weighted")) |>
  mutate(dAIC = AIC_no_sonar - AIC_rl_coarse,
         dLL =  LL_rl_coarse - LL_no_sonar)

gf_violin(dLL ~ weighted, data = sim_results2) |>
  gf_sina() |>
  gf_refine(scale_y_continuous(trans='log2')) |>
  gf_labs(y = "AIC difference\n(with minus without MFAS covariate)",
          x = "Weighted Likelihood?")
```

By what margin would model selection using AIC select the true model (*without* MFAS effect -- so a small AIC difference would be hoped for), with and without duration-weighting applied to the likelihood?
:::

\newpage

::: {#supptab-daic-with-weights2}

```{r}
selection_table2 <- mosaic::prop(~dAIC >= 6 | weighted, data = sim_results2) |>
  data.frame()
names(selection_table2) <- "Proportion dAIC > 6"

selection_table2 <- selection_table2 |>
  mutate(`Weighting?` = c("No", "Yes")) |>
  select("Weighting?", "Proportion dAIC > 6")
  
  gt(selection_table2)
```

Proportion of simulations where the difference in AIC (model with MFAS minus model without) exceeded 6, when the true model was one *without* MFAS effect.
:::

#### Confidence Interval Coverage

```{r}
sim_CI2 <- sim_CI2 |>
  mutate(true_15 = 0,
         true_51 = 0 ) |>
  mutate(true_15_in_CI = if_else(t15lower <= true_15 &
                                   t15upper >= true_15,
                                 1,
                                 0),
         true_51_in_CI = if_else(t51lower <= true_51 &
                                   t51upper >= true_51,
                                 1,
                                 0),
         row_num = c(1:n()),
         pretty_name = case_when(model_name == "rl_coarse" ~ "No Weighting",
                                 model_name == "rl_coarse_wt" ~ "Weighted"))

CI_summary2 <- sim_CI2 |>
  drop_na(t15est) |>
  group_by(model_name) |>
  summarise(
    # n_NA_15 = sum(is.na(true_15_in_CI)),
    # n_NA_51 = sum(is.na(true_51_in_CI)),
    n_total = n(),
    prop_in_CI_15 = sum(true_15_in_CI, na.rm = TRUE) / sum(!is.na(true_15_in_CI)) * 100,
    prop_in_CI_51 = sum(true_51_in_CI, na.rm = TRUE) / sum(!is.na(true_51_in_CI)) * 100) |>
  ungroup() |>
  mutate(across((starts_with("MAE") | starts_with("prop_")), function(x) signif(x, 3))) |>
  rename(`Model Name` = model_name,
         # `NAs in P(T->V)` = n_NA_15,
         # `NAs in P(V->T)` = n_NA_51, # very hard to estimate b/c sample size too small
         `N Simulations` = n_total,
         `CI Coverage (%; P(T->V))` = prop_in_CI_15,
         `CI Coverage (%; P(V->T))` = prop_in_CI_51
         )
```

::: {#supptab-ci-summary2}

```{r}
gt(CI_summary2) |>
  tab_style(
    style = cell_text(size = pct(65)),
    locations = list(cells_body(), cells_column_labels()) 
  )
```

Confidence interval coverage in simulations, for simulations in which the true model *did not include* an effect of MFAS on the transition rates. "T->V" is the probability of transition from Typical to Variant state, and "V->T" is Variant to Typical.
:::

# Discussion

Here, we briefly reported results from a simulation study that arose during our attempts to identify a workable approach to fit the HHMMs: assessing the impact of applying likelihood weights according to time-step duration, motivated by that fact that fine-scale observations may outnumber coarse-scale observations by an order of magnitude. 

While with hindsight it is unsurprising that such a weighting scheme biases inference by yielding unrealistically low uncertainty at the coarser timescale, we still present the finding here as a reminder that, while weighting can be part of a highly effective strategy to help tune *predictive* models (as seen in Sidrow *et al.* 2025 (<https://doi.org/10.1371/journal.pone.0325321>)), a distinct modeling process may be required for accurate *inference* (as pointed out by Bzdok and Ioannidis 2019 (<https://doi.org/10.1016/j.tins.2019.02.001>) and Tredennick *et al.* 2021 (<https://doi.org/10.1002/ecy.3336>)).